---
title: "Final_Draft"
author: "Stephanie Spencer"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)


data <- read.csv("UtahFireflies_Revised.csv")
```

# ABSTRACT

# BACKGROUND
Citizens have the power to help scientific observations move along. The widespread ability to collect data is amazing. Using their observations of fireflies across Utah we have the opportunity to monitor the firefly population across the years. This data set includes retroactive data from before 2000, however, by focusing on the information from 2016 to 2024, we can determine more recent trends in the firefly population. In Evans et al. paper from 2019 they found a link between the larval stage abundance and the climate, where the temperature in the prevoius year would affect the abundance and timing of the next year's larval stage. This is an indication that temeprature could be having an effect on the firefly abundance each year. 

# QUESTIONS AND HYPOTHESIS

## Question
How are the sightings of fireflies and their abundance impacted by month and yearly conditions?

## Hypothesis
We will see a decrease in fireflies each year due to global warming or other temperature affects. 
## Prediction
We will see a common trend of when firefly activity/population peaks each month. We will see limited variation by year as their activity may not be reactive to the weather trends. 
# METHODS
Using a data set of firefly observations from Logan's citizen science data, we could extract sighting dates, locations, and habitat type. The number sighted were concluded to be inconsistent results as many were estimated numbers and could be very easily skewed by perspective. Therefore, in this study we focused on the instance report. Counting how many times per month/year fireflies were observed. Using the ggplot2 package in R would could view how these trends differ by month over years. To zoom in more, we can use Julian days of the peak observation day of each year to show if the fireflies are peaking earlier or later. This data can be compared with climate/weather data. To quantify how different each year was, and if there was a linear trend we used linear modeling and ANOVA test for signifigance. 

## Figure 1
```{r , warning= FALSE}
ggplot(data, aes(x = Month)) +
  geom_bar(fill = "yellow3") +
  facet_grid(~ Year) +
  labs(title = "Observations Recorded by Month and Year",
       x = "Month",
       y = "Number of Observations") +
  theme_minimal()
#Edited with AI to implement/ fix these functions.
```



## Figure 2
```{r maximum Julian day, warning=FALSE}

#make plot of Max Julian day
table_Julian <- table(data$Year, data$Julian)
#find the max count each year
max_col <- apply(table_Julian, 1, which.max) 
# create a data frame with year and max Julian day
Max_Julian_Day <- as.numeric(colnames(table_Julian)[max_col])
years <- c(2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024)
#plot as scatter plot
plot(years, Max_Julian_Day, main = "Peak Days of Observations", xlab = "Year", ylab = "Julain Days")

```

## Figure 3
```{r linear model}
#create data frame 
Julian_dataframe <- data.frame(ColumnA = Max_Julian_Day, ColumnB = years)
m1 <- lm( ColumnA ~ ColumnB, data=Julian_dataframe)
# run anova test for p value
anova(m1)
summary(m1)
```

# DISCUSSION

## Analysis of Figure 1
We see a consistency across the 9 years that firefly observations peak in June each year. The size of the peaks varies,  with a seemingly decreased trend of observations. However, as this data come from citizen science many factors could impact the numbers including how often people continue to submit sightings after several years.

## Analysis of Figure 2


## Analysis of Figure 3

# CONCLUSION

# REFERENCES

Citizen Science Libraries. (n.d.). UtahFireflies. Google Docs. Retrieved October 30, 2025, from https://docs.google.com/spreadsheets/d/1XLklQd6P3d9wYnmh-3KgzNvP8GfQHL2Y2IyaZMwIOzk/edit?usp=drive_web&ouid=106039106225729317949&usp=embed_facebook

Evans, T. R., Salvatore, D., van de Pol, M., & Musters, C. j. m. (2019). Adult firefly abundance is linked to weather during the larval stage in the previous year. Ecological Entomology, 44(2), 265–273. https://doi.org/10.1111/een.12702

OpenAI. (2025, October 30). Barplot creation in R [Generative AI chat]. ChatGPT.


